---
title: "Mapping Parameters Evaluation"
author: "Nastassia Gobet"
date: '`r Sys.Date()`'
output:
  html_document:
    keep_md: no
    code_folding: hide
    fig_caption: yes
    highlight: tango
    number_sections: no
    theme: readable
    toc: yes
    toc_depth: 4
    df_print: paged
  pdf_document:
    fig_caption: yes
  word_document:
    toc: yes
    toc_depth: '4'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path="Graphs/MappingParametersEvaluation/", warning=TRUE)

#load needed libraries
library(knitr)
library(vioplot)

# set working directory
##setwd("F:/BXD/analysis/Mapping")
```

# GOAL

Determine the best mapping settings and compare BXD-specific references to B6 reference.

# Analysis

## Data preparation

Extract number of significant eQTLs and total eQTLs.

```{r RetrievingeQTLsStats, fig.height=8, fig.align="center"}
basepath <- "F:/BXD/data/MappingEvaluation/"
settinglist <- list.dirs(path=basepath, full.names=FALSE, recursive=FALSE)
tissuelist <- c("Cortex", "Liver")
conditionlist <- c("NSD", "SD")
##normalizationtypelist <- c("log2CPM", "log2TPM")
normalizationtypelist <- c("log2CPM")

# function to retrieve number of eQTLs and number of expressed genes
retrieveeQTLstats <- function(basepath, setting, normalizationtype, tissue, condition){
  # load eQTL data
  ##print(paste0(basepath, setting, "/TMMnormalized_", normalizationtype, "_", tissue, "_", condition, "_pvalcorrected.txt"))
  data <- read.table(paste0(basepath, setting, "/TMMnormalized_", normalizationtype, "_", tissue, "_", condition, "_pvalcorrected.txt"), stringsAsFactors=FALSE)
  # retrieve number of significant eQTLs
  sig <- length(which(data$adjustedpvalue<0.05))
  # retrieve total number of eQTLs
  tot <- nrow(data)
  
  return(c(nb_sig=sig, total=tot, ratio=sig/tot))
}

##retrieveeQTLstats(basepath, setting, normalizationtype, tissue, condition)

# determine best (= more significant eQTLs) mapping setting for each combination of tissue, condition, and type of normalization
idx <- 1
optimalsettings <- as.data.frame(matrix(NA, nrow=8, ncol=7))
colnames(optimalsettings) <- c("Condition", "Tissue", "Type of Normalization", "# significant eQTLs", "# expressed genes", "Percentage of significant eQTLs", "Setting")
for(tissue in tissuelist){
  for(condition in conditionlist){
    for(normalizationtype in normalizationtypelist){
      optimalsettings[idx, "Tissue"] <- tissue
      optimalsettings[idx, "Condition"] <- condition
      optimalsettings[idx, "Type of Normalization"] <- normalizationtype
      eQTLstats <- as.data.frame(t(sapply(settinglist, retrieveeQTLstats, basepath=basepath, normalizationtype=normalizationtype, tissue=tissue, condition=condition)))
      sort_idx <- sort(eQTLstats$ratio, index.return=TRUE)$ix
      dotchart(eQTLstats$ratio[sort_idx]*100, labels=gsub("genotypesandimputed_", "", rownames(eQTLstats)[sort_idx]), cex=0.8, main=paste(tissue, condition, normalizationtype), pch=19, col=c("darkmagenta","orchid"), xlab="Percentage of significant eQTLs")
      gsub("genotypesandimputed_", "", rownames(eQTLstats))
      gsub("genotypesandimputed_", "", rownames(eQTLstats))
      bestsetting <- eQTLstats[which.max(eQTLstats$ratio),]
      optimalsettings[idx, "Percentage of significant eQTLs"] <- bestsetting$ratio*100
      optimalsettings[idx, "# significant eQTLs"] <- bestsetting$nb_sig
      optimalsettings[idx, "# expressed genes"] <- bestsetting$total
      optimalsettings[idx, "Setting"] <- rownames(bestsetting)
      idx <- idx + 1
    }
  }
}
```

### Mapping evaluation results

```{r MappingEvaluation}
# display results as table
optimalsettings
```

### Merging results for conditions

To have one unique classification of best mapping parameters per organ take the average between conditions (NSD and SD).

```{r MergingMappingParametersResults, fig.height=7, fig.width=5}
# retrieve all values for log2CPM data
normalizationtype <- "log2CPM"
for(tissue in tissuelist){
  for(condition in conditionlist){
      assign(paste("eQTLstats", tissue, condition, sep="_"), as.data.frame(t(sapply(settinglist, retrieveeQTLstats, basepath=basepath, normalizationtype=normalizationtype, tissue=tissue, condition=condition))))
  }
}

averageratio_Cortex <- apply(cbind(eQTLstats_Cortex_NSD$ratio, eQTLstats_Cortex_SD$ratio), MARGIN=1, FUN=mean)
order_Cortex <- sort(averageratio_Cortex, index.return=TRUE)$ix

averageratio_Liver <- apply(cbind(eQTLstats_Liver_NSD$ratio, eQTLstats_Liver_SD$ratio), MARGIN=1, FUN=mean)
order_Liver <- sort(averageratio_Liver, index.return=TRUE)$ix

dotchart(averageratio_Cortex[order_Cortex]*100, labels=gsub("genotypesandimputed_", "", rownames(eQTLstats_Cortex_NSD)[order_Cortex]), cex=0.8, main=paste("Cortex", normalizationtype), pch=19, col=c("darkmagenta","orchid"), xlab="significant eQTLs (%)")

dotchart(averageratio_Liver[order_Liver]*100, labels=gsub("genotypesandimputed_", "", rownames(eQTLstats_Liver_NSD)[order_Liver]), cex=0.8, main=paste("Liver", normalizationtype), pch=19, col=c("darkmagenta","orchid"), xlab="significant eQTLs (%)")
```

### Alternative measures

Alternative measures as objective function to evaluate to try to avoid reference bias.

An eQTL with a negative slope means that the B6 expression is higher than D2.


# Session information

```{r}
sessionInfo()
```
